---
title: "A Quick Start for SpaXTalk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Quick Start for SpaXTalk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The examples are not evaluated while the package vignette is built because they
require the external SpaXTalk sample dataset. After downloading the dataset,
you can run the examples interactively.

## Step 1: Load required packages

```{r load-packages}
library(SpaXTalk)
library(dplyr)
library(Matrix)
```

## Step 2: Load the spatial transcriptomics data

An example dataset of mouse embryo Stereo-seq data is available
[here](https://drive.google.com/file/d/1Xo40BXmw7u1EVjvJ2RlYh30Xrky8mSAF/view?usp=drive_link).

This dataset is provided as a Seurat object. For more information about Seurat,
see the [Seurat documentation](https://satijalab.org/seurat/articles/get_started_v5_new).

Load the downloaded data using the following code:

```{r load-data}
SeuratObj <- readRDS("E:/SpaXTalk_SampleData.RDS")
```

## Step 3: Prepare the SpaXTalk input

SpaXTalk requires a gene-by-cell expression matrix and cell metadata containing
x coordinates, y coordinates, and cell annotations. By default, the tutorial
assumes these values are stored in the `x`, `y`, and `annotation` metadata
columns, respectively.

```{r prepare-input-manually}
exp_data <- SeuratObject::GetAssayData(
  SeuratObj,
  assay = "RNA",
  layer = "data"
)

meta_all <- SeuratObj@meta.data
meta <- data.frame(
  x = meta_all$x,
  y = meta_all$y,
  annotation = as.character(meta_all$annotation),
  row.names = rownames(meta_all),
  stringsAsFactors = FALSE
)

# Specify the receiver population of interest.
receiver <- rownames(meta)[meta$annotation %in% "Neural_crest"]

# Release the memory
rm(SeuratObj)
```

Alternatively, for data stored in a Seurat object, `Prepare_data()` can generate the required
inputs directly:

```{r prepare-input-helper}
prepared <- Prepare_data(
  SeuratObj,
  receiver_cluster = "Neural_crest",
  assay_type = "RNA",
  layer_type = "data"
)

exp_data <- prepared$exp_data
meta <- prepared$meta
receiver <- prepared$receiver

# Release the memory
rm(SeuratObj)
```

`Prepare_data()` returns the expression matrix, standardized metadata, and
receiver names in a named list.

## Step 4: Load the prior knowledge

SpaXTalk provides a built-in cell-cell communication database extracted from
CellChat and an intracellular gene-regulation database derived from KEGG. Load
the two knowledge bases into the current workspace:

```{r load-databases}
data("CellChat_mouse", package = "SpaXTalk") # Load LRDB
data("KEGG_PPR_mouse", package = "SpaXTalk") # Load RTDB
```

## Step 5: Perform the SpaXTalk analysis

```{r run-analysis}
neighbor_list <- Identify_neighbor_list(
  locs = meta[, c("x", "y"), drop = FALSE],
  receivers = receiver,
  dist_threshold = 2.5,
  decay_type = "exp",
  normalize = TRUE,
  nclusters = 1
)

neighbor_list[[1]] # Inspect one neighborhood.

CCC_origin <- InferCCC(
  data = exp_data,
  neighbor_list = neighbor_list,
  LRDB = LRDB,
  nclusters = 1
)

sum(CCC_origin[[1]])

GRNs <- InferGRN(
  exps = exp_data,
  neighbor_list = neighbor_list,
  RTDB = RTDB,
  PPR_top = 5,
  exp_thres = 0.25,
  nsim = 0,
  lambda = 1,
  nclusters = 16,
  use_delta_y = TRUE
)

betas <- GRNs$beta

# Save the results (optional).
allresults <- list(
  CCC = CCC_origin,
  GRN = GRNs
)
saveRDS(allresults, "SpaXTalk_results.rds")
```

## Step 6: Process the results

```{r process-results}
# Calculate the crosstalk.
CCC_mat <- Aggr_CCC_results(CCC_origin, flag = "Sender")
CCC_df <- tidyfst::mat_df(CCC_mat)
GRN_filtered <- Process_GRN(allresults$GRN, beta_thres = 1e-5)
XT_results <- Aggr_XT_results(CCC_df, GRN_filtered, threshold = 1e-8)
```

## Step 7: Visualize the results

### Plot cell-cell communication

Plot the cluster-level communication pattern:

```{r plot-dominant-lr}
Plot_DominantLR(results = CCC_mat)
```

Plot communication in space:

```{r plot-spatial-communication}
data <- Aggr_CCC_results(CCC_origin, flag = "LR", LR_used = NULL)
PlotSpa_SRVector(meta = meta, data = data, threshold = 1)

# Plot individual ligand-receptor pairs.
features <- c("Mif=Cd44", "Mdk=Ncl", "Mdk=Itgb1", "Igf2=Igf2r")

data <- Aggr_CCC_results(
  CCC_origin,
  flag = "LR",
  LR_used = features[1]
)
PlotSpa_SRVector(meta = meta, data = data, threshold = 0.75)

data <- Aggr_CCC_results(
  CCC_origin,
  flag = "LR",
  LR_used = features[2]
)
PlotSpa_SRVector(meta = meta, data = data, threshold = 0.75)
```

### Plot crosstalk

Crosstalk can be visualized at either the single-cell or cell-group level.
Plot the crosstalk pattern of one cell:

```{r plot-single-cell-crosstalk}
color_use <- c(
  Ligand = "aquamarine",
  Receptor = "bisque",
  Target = "pink"
)

XT_filtered <- XT_results %>%
  filter(Cell == "185_151") %>%
  arrange(desc(Weight)) %>%
  slice_head(n = 6)

Plot_MultiLayer_Singlecell(
  XT_filtered,
  start_trim = 0.3,
  end_trim = 0.3,
  arrow_len_cm = 0.25,
  x_scale = 3.25,
  rect_width = 3,
  rect_height = 0.3,
  label_size = 6,
  layer_fill_colors = color_use
)
```

Plot crosstalk for the whole cell group:

```{r plot-group-crosstalk}
aggr_XT <- aggregate(
  Weight ~ Ligand + Receptor + TF,
  data = XT_results,
  sum
)

PlotXT_Alluvial(
  aggr_XT,
  features = NULL,
  topk = 20,
  vertical = FALSE
)
```

Plot the spatial distribution of the specificity of a single transcription
factor:

```{r plot-spatial-specificity}
tf <- XT_results$TF[which.max(XT_results$Weight)]

aggr_spe <- XT_results %>%
  group_by(Cell, TF) %>%
  summarise(Weight = sum(Weight), .groups = "drop_last") %>%
  mutate(Spe = Weight / sum(Weight)) %>%
  ungroup() %>%
  filter(TF == tf)

specificity <- setNames(aggr_spe$Spe, aggr_spe$Cell)

PlotSpa_Scatter(
  specificity,
  meta,
  legend_name = paste(tf, "specificity")
)
```
